---
title: "Assignment 2: Wrapping up regression"
author: 
  - "Lydia Hefel"
  - "DS303, SP26"
  - "Prof. Amber Camp"
date: today
format: html
toc: true
editor: visual
theme: spacelab
---

# Wrapping up linear and logistic regression^\*^

Assignment 2 covers linear and logistic regression models and includes many of the topics we have covered over the entire semester. You will be using `palmerpenguins` data, which includes a nice variety of continuous and categorical predictors.

To submit this assignment, render this file and save as a pdf. Commit & push this qmd to GitHub and upload the pdf to Canvas.

\*This assignment does not include mixed effects models, but you will be seeing that on another assignment.

## Load Packages

```{r, message = FALSE}
library(lme4)
library(tidyverse)
# install.packages("palmerpenguins") # install if needed
library(palmerpenguins)
```

## Load Data

Load the `penguins` data and examine it below (use `summary()`, etc.)

```{r}
penguins <- penguins

head(penguins)
summary(penguins)
??penguins
```

## Question 1: Describe the data

What information is contained in this data set? Describe at least four variables (excluding `year`), including what they represent and their data type. Lastly, describe whether you think `year` would be a useful predictor in this data.

**Answer**:\
`species` - This variable represents the specie of penguin that was recorded (Adelie, Chinstrap, Gentoo) and the data type is factor.

`bill_len` - This variable represents the length, in millimeters, of an individual penguin's bill and the data type is numeric.

`bill_dep` - This variable represents the depth, in millimeters, of an individual penguin's bill and the data type is numeric.

`flipper_len` - This variable represents the length, in millimeters, of an individual penguin's flippers and the data type is integer/numeric.

`year` - I think `year` has the potential to be a useful predictor in this data, especially if there was any sort of mass migration or extinction of a penguin specie in any of the years (2007-2009).

## Question 2: EDA

Explore your data visually. Create at least two visualizations that show the relationship between `flipper_length_mm` and its potential predictors.

```{r, warning = FALSE}
ggplot(penguins, aes(x = flipper_length_mm, y = species, color = species)) +
  geom_jitter()

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_jitter()

ggplot(penguins, aes(x = flipper_length_mm, y = sex, color = species)) +
  geom_jitter()

ggplot(penguins, aes(x = flipper_length_mm, y = as.factor(year), color = species)) +
  geom_jitter()

ggplot(penguins, aes(x = flipper_length_mm, y = island, color = species)) +
  geom_jitter()

ggplot(penguins, aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_jitter()

ggplot(penguins, aes(x = flipper_length_mm, y = bill_depth_mm, color = species)) +
  geom_jitter()
```

## Question 3: Apply a linear regression

Fit a simple linear regression model predicting `flipper_length_mm` from `body_mass_g`.

```{r}
simple <- lm(flipper_length_mm ~ body_mass_g, data = penguins)
summary(simple)
```

Interpret your model output, including the intercept and slope, in your own words below. Be sure to include a sentence explaining how `body_mass_g` impacts `flipper_length_mm` and whether or not the effect is significant.

**Answer:**

-   **Intercept**: 1.367e+02

    -   The `flipper_length_mm` is 1.367e+02 units when `body_mass_g` is zero.

-   `body_mass_g`: 1.528e-02

    -   For every one unit increase in `body_mass_g`, there is a *statistically significant* increase in `flipper_length_mm` by 1.528e-02 units.

## Question 4: Apply a multiple linear regression

Fit a linear regression model predicting `flipper_length_mm` from both `body_mass_g` and `bill_length_mm`. Interpret the slopes and intercept.

```{r}
multiple <- lm(flipper_length_mm ~ body_mass_g + bill_length_mm, data = penguins)
summary(multiple)
```

Similar to Question 3, interpret the model output in your own words here:

**Answer:**

-   **Intercept**: 1.220e+02

    -   The `flipper_length_mm` is 1.220e+02 units when all other variables are zero.

<!-- -->

-   `body_mass_g`: 1.305e-02

    -   For every one unit increase in `body_mass_g`, there is a *statistically significant* increase in `flipper_length_mm` by 1.305e-02 units, while `bill_length_mm` is held constant.

-   `bill_length_mm`: 5.492e-01

    -   For every one unit increase in `bill_length_mm`, there is a *statistically significant* increase in `flipper_length_mm` by 5.492e-01 units, while `body_mass_g` is held constant.

## Question 5: Include an interaction

Fit a linear regression model predicting `flipper_length_mm` from `body_mass_g,` `bill_length_mm`, and the **interaction** of the two. Interpret the slopes and intercept.

```{r}
interaction <- lm(flipper_length_mm ~ body_mass_g * bill_length_mm, data = penguins)
summary(interaction)
```

Interpret the model output in your own words below. If there was a change in the pattern of significance, try to explain the logic below as well.

**Answer:**

-   **Intercept**: 1.706e+02

    -   The `flipper_length_mm` is 1.706e+02 units when all other variables are zero.

<!-- -->

-   `body_mass_g`: 4.364e-04

    -   For every one unit increase in `body_mass_g`, there is an increase in `flipper_length_mm` by 4.364e-04 units, while `bill_length_mm` is held constant.

-   `bill_length_mm`: -5.051e-01

    -   For every one unit increase in `bill_length_mm`, there is a decrease in `flipper_length_mm` by 5.051e-01 units, while `body_mass_g` is held constant.

-   `body_mass_g:bill_length_mm`: 2.707e-04

    -   When `body_mass_g` increases by one unit, the effect of `bill_length_mm` on `flipper_length_mm` shifts by 2.707e-04 units.

-   There was a change in the pattern of significance as `body_mass_g` and `bill_length_mm` are no longer statistically significant. Their interaction is much more significant, thus their individual effects are no longer as significant.

## Question 6: Compare models

Compare the models you built in Questions 4 and 5 using `anova()`.

```{r}
anova(multiple, interaction)

anova(simple, interaction)
```

-   Which is the better model? How do you know?

    -   **Answer:**
        -   The model from Question 5, with the interaction, is the better model because it has a lower RSS in comparison to the model from Question 4.

-   Is it possible to compare the models from Questions 3 and 5 using the same method? Why or why not?

    -   **Answer:**
        -   Yes, it is possible to compare the models from Questions 3 and 5 using the same method. Since ANOVA compares models hierarchical, and the model from Question 3 is nested in the model from Question 5, the two models are able to be compared (and the model from Question 5 has a lower RSS so it is better in comparison).

## Question 7: Categorical predictors

Build a linear model that includes a categorical predictor of your choice. It is fine to stick with dummy coding. Optional: apply a different coding scheme AND interpret the output correctly for +1 extra credit.

```{r}
contrasts(penguins$species)

dummy <- lm(flipper_length_mm ~ species, data = penguins)
summary(dummy)
```

-   What is the reference level of your categorical predictor?

    -   **Answer:**
        -   The reference level of species is Adelie because, in this case, it comes first alphabetically.

-   What is your interpretation of this model output? Address all coefficients.

    -   **Answer:**
        -   The average `flipper_length_mm` of the Adelie penguins is 189.9536 units.
        -   Compared to the Adelie penguins, the Chinstrap penguins `flipper_length_mm` increases by 5.8699 units, this coefficient is statistically significant.
        -   Compared to the Adelie penguins, the Gentoo penguins `flipper_length_mm` increases by 27.2333 units, this coefficient is statistically significant.

## Question 8: Relevel your categorical variable

Relevel your categorical variable so that a **different** level becomes the reference. Then, run the same model you did in Question 7 and interpret the output.

Relevel:

```{r}
penguins$species <- relevel(penguins$species, ref = "Gentoo")
```

Apply same model from Question 7:

```{r}
contrasts(penguins$species)

relevel <- lm(flipper_length_mm ~ species, data = penguins)
summary(relevel)
```

-   What is the new reference level of your categorical predictor?

    -   **Answer:**
        -   The reference level of species is now Gentoo.

-   What is your interpretation of this new model output? Address all coefficients.

    -   **Answer:**
        -   The average `flipper_length_mm` of the Gentoo penguins is 217.1870 units.
        -   Compared to the Gentoo penguins, the Chinstrap penguins `flipper_length_mm` decreases by 27.2333 units, this coefficient is statistically significant.
        -   Compared to the Gentoo penguins, the Chinstrap penguins `flipper_length_mm` decreases by 21.3635 units, this coefficient is statistically significant.

## Question 9: Fit a logistic regression

Apply a logistic regression. Include as many predictor variables as you'd like. Remember that your predicted outcome variable needs to be binary (or categorical with two levels).

Hint: You could use `sex` or create a binary variable of your own (e.g., Gentoo vs. non-Gentoo) to test your model.

```{r}
logistic <- glm(sex ~ flipper_length_mm, data = penguins, family = binomial)
summary(logistic) 
```

What are your key takeaways from this model?

**Answer:**

-   **Intercept**: -7.65801

    -   This is the log-odds of being classified as `female` when `flipper_length_mm` is zero.

-   `flipper_length_mm`: 0.03822

    -   For each one-unit increase in `flipper_length_mm`, the log-odds of a penguin being classified as `female` (rather than `male`) decreases by 0.03822 units, this coefficient is statistically significant.

    -   Since the coefficient is positive, it suggests that as `flipper_length_mm` increases, the log-odds of being classified as `female` significantly decreases, compared to `male`.

## Question 10: Synthesize the information

Imagine you’re a biologist studying penguin populations. Which predictors do you think are most important to measure or record in the field to predict flipper length? Why?

**Answer:**

-   Imagining myself as a biologist studying penguin populations, the most important predictors to measure or record in the field to predict flipper length are body mass, bill depth, species, and island. The EDA from Question 2 showed that these variables led to the most distinguishable outputs, and are able to help determine flipper length. A record of these predictors, and the interactions among these predictors, could help create better models that can more accurately predict flipper length.

## Bonus: Stepwise Regression

Perform stepwise regression to find the best model for an outcome of your choice. You will likely encounter an error about missing values. Fixing that error and explaining your findings will earn you +2 extra credit. Show your work.

```{r}


```

According to this stepwise regression, explain how the final model was selected.

**Answer:**\
(Write answer here)

## Add Mixed Effects Models

```{r}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

### Fixed Effects Only Model

```{r}
m_fixed <- lm(flipper_length_mm ~ body_mass_g + species, data = penguins)
summary(m_fixed)
```

### Random Intercepts Model

```{r}
library(lmerTest)

m_ri <- lmer(flipper_length_mm ~ body_mass_g + (1 | species), data = penguins)
summary(m_ri)

ranef(m_ri)
```

### Random Intercepts + Random Slopes

```{r}
m_ris <- lmer(flipper_length_mm ~ body_mass_g + (1 + body_mass_g | species), data = penguins)
summary(m_ris)

ranef(m_ris)
```

### Model Comparison

```{r}
anova(m_ri, m_ris)
#anova(m_fixed, m_ri) - cannot compare fixed and mixed effect models with anova

AIC(m_fixed, m_ri)
```

## Revisit Sleep Study

```{r}
sleepstudy <- sleepstudy

ggplot(sleepstudy, aes(x = Days, y = Reaction)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()

sleep_fixed <- lm(Reaction ~ Days, data = sleepstudy)
summary(sleep_fixed)

sleep_mixed <- lmer(Reaction ~ Days + (1 | Subject), data = sleepstudy)
summary(sleep_mixed)

AIC(sleep_fixed, sleep_mixed)
```

### R-Squared

```{r}
#install.packages("MuMIn")
library(MuMIn)
r.squaredGLMM(sleep_mixed)
```
