---
title: "Risk Factors as Predictors of Cervical Cancer"
subtitle: "Prediction of Cervical Cancer Indicators"
author: "Lydia Hefel"
institute: "Chaminade University"
date: 5/4/25
format: 
  revealjs:
    theme: dracula
editor: visual
---

## Background

::: incremental
-   \~11, 000 new cases a year in the U.S.

-   Number of new cases declining over the years

-   Takes 4,000 U.S. lives and 300,000 lives worldwide each year

-   Labeled the most preventable cancer

-   Screenings (i.e. pap smears) can detect pre-cancerous cells
:::

## Research Question

Which risk factors are the best predictors of whether or not a person is diagnosed with cervical cancer?

<br>

. . .

**Why it matters...**

. . .

-   Most preventable cancer yet 300,000 deaths worldwide each year

-   Understand high risk lifestyle choices

-   Improve detection process

-   Better prevention

## Dataset {.smaller}

-   Found on **Kaggle** - *Cervical Cancer Risk Classification*

-   Sourced from **UCI ML Repository** - *Cervical Cancer Risk Factors for Biopsy*

-   Collected at **Hospital Universitario de Caracas** in *Caracas, Venezuela*

<br>

. . .

***Why this dataset?***

. . .

-   Risk factor classification

-   Publicly available with multiple copies

<br>

. . .

***Key Variables***

. . .

-   Hypothesis: `num_of_pregnancies` and `first_sexual_intercourse`

-   Results: `dx_hpv`, `smokes`, and `num_of_pregnancies`

## Data Cleaning Process {.smaller}

<br>

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(plotly)
library(patchwork)
library(here)
library(janitor)
library(dplyr)
library(plotly)
library(ISLR2)
```

```{r, echo=TRUE, output=FALSE}
project_data <- read_csv(here("data/kag_risk_factors_cervical_cancer.csv"))

project_data <- clean_names(project_data)

project_data <- project_data %>%
  rename_with(~ gsub("st_ds", "stds", .))

colSums(project_data == "?", na.rm = TRUE)

project_data <- project_data %>%
  mutate(across(where(is.character), ~na_if(.,"?")))

project_data <- project_data %>%
  mutate(across(where(is.numeric), as.factor))

project_data <- project_data %>%
  mutate(across(where(is.character), as.numeric))

project_data <- project_data %>%
  mutate(across(c(iud, stds, smokes, hormonal_contraceptives), as.factor))

project_data <- project_data %>%
  mutate(across(c(age), as.numeric))

#stds_time_since_first_diagnosis and stds_time_since_last_diagnosis only ~8% complete
```

## Exploratory Data Analysis I

<br>

```{r}
ggplotly(ggplot(project_data, aes(x = dx_cancer, y = num_of_pregnancies, color = first_sexual_intercourse)) +
  geom_jitter(na.rm = TRUE) +
  theme_minimal() +
  labs(title = "Cervical Cancer Diagnoses by Number of Pregnancies and Age of First Sexual Intercourse",
       x = "Cervical Cancer Diagnosis",
       y = "Number of Pregnancies",
       color = "Age of First Sexual\n      Intercourse"))
```

## Exploratory Data Analysis II

<br>

```{r}
ggplotly(ggplot(project_data, aes(x = dx_cancer, y = smokes, color = smokes_packs_year)) +
  geom_jitter(na.rm = TRUE) +
  theme_minimal() +
  labs(title = "Cervical Cancer Diagnoses by Patient Smoking Frequency",
       x = "Cervical Cancer Diagnosis",
       y = "Patient Smokes",
       color = "Average Number of Packs\n   Patient Smokes Yearly"))
```

## Exploratory Data Analysis III

<br>

```{r}
ggplotly(ggplot(project_data, aes(x = dx_cancer, y = dx_hpv, color = age)) +
  geom_jitter(na.rm = TRUE) +
  theme_minimal() +
  labs(title = "Cervical Cancer Diagnoses by HPV Diagnoses and Patient Age",
       x = "Cervical Cancer Diagnosis",
       y = "HPV Diagnosis",
       color = "Patient Age"))
```

## Logistic Regression Model I {.smaller}

<br>

```{r}
vars <- c("dx_cancer", "dx_hpv", "num_of_pregnancies", "smokes")
clean_data <- project_data[complete.cases(project_data[, vars]), ]
```

```{r, echo=TRUE, output=FALSE}
hpv_diagnosis <- glm(dx_cancer ~ dx_hpv, clean_data, family = binomial)
```

```{r, echo=FALSE, output=TRUE}
hpv_diagnosis <- glm(dx_cancer ~ dx_hpv, clean_data, family = binomial) 
summary(hpv_diagnosis) 
```

## Logistic Regression Model II {.smaller}

<br>

```{r, echo=TRUE, output=FALSE}
hpv_pregnancies <- glm(dx_cancer ~ dx_hpv : num_of_pregnancies, clean_data, family = binomial) 
```

```{r, echo=FALSE, output=TRUE}
hpv_pregnancies <- glm(dx_cancer ~ dx_hpv : num_of_pregnancies, clean_data, family = binomial) 
summary(hpv_pregnancies)
```

## Logistic Regression Model III {.smaller}

<br>

```{r, echo=TRUE, output=FALSE}
hpv_smoking <- glm(dx_cancer ~ dx_hpv : smokes, clean_data, family = binomial)
```

```{r, echo=FALSE, output=TRUE}
hpv_smoking <- glm(dx_cancer ~ dx_hpv : smokes, clean_data, family = binomial)
summary(hpv_smoking)
```

## Logistic Regression Model IV {.smaller}

<br>

```{r, echo=TRUE, output=FALSE}
combined <- glm(dx_cancer ~ dx_hpv : num_of_pregnancies : smokes, clean_data, family = binomial)
```

```{r, echo=FALSE, output=TRUE}
all <- glm(dx_cancer ~ dx_hpv : num_of_pregnancies : smokes, clean_data, family = binomial)
summary(all)
```

## ANOVA Model Comparison

::::: columns
::: {.column width="50%"}
```{r, echo=TRUE}
anova(hpv_diagnosis, hpv_pregnancies, all)
```
:::

::: {.column width="50%"}
```{r, echo=TRUE}
anova(hpv_diagnosis, hpv_smoking, all)
```
:::
:::::

## AIC Model Comparison

<br>

```{r, echo=TRUE}
AIC(hpv_diagnosis, hpv_pregnancies, hpv_smoking, all)
```

## Findings

<br>

. . .

HPV diagnosis was the most statistically significant predictor of cervical cancer.

<br>

. . .

Number of pregnancies and smoking frequency were not significant on their own...

. . .

but they became significant when interacting with HPV diagnosis.

## Limitations

::: incremental
-   Dataset size

-   Variable definitions

-   Demographic representation
:::

<br>

. . .

### **Ethical Considerations**

::: incremental
-   Patient privacy

-   Consent to patient data
:::

# Thank you!
